The Impacts of Aquifer Depletion on the Ability of Groundwater Irrigation to Buffer Against Undesirable Weather

### 2022-09-03 Writing/manuscript_DR.qmd
initial commit

---
title: The Impacts of Aquifer Depletion on the Ability of Groundwater Irrigation to Buffer Against Undesirable Weather
author:
  - Taro Mieno^[Department of Agricultural Economics, University of Nebraska Lincoln, tmieno2@unl.edu]
  - Timothy Foster^[Department of WWW, University of Manchester, timothy.foster@manchester.ac.uk]
  - Shunkei Kakimoto^[Department of Applied Economics, University of Minnesota]
  - Nicholas Brozovic^[Department of Agricultural Economics, University of Nebraska Lincoln, nbrozovic@nebraska.edu]
abstract: |
  Abstract goes here. It should be a single paragraph of around 200 words.
acknowledgments: |
  We wish to thank...
bibliography: DRA.bib
csl: iopart.csl
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: true
    toc: false
---

@@

#--- packages ---#
library(data.table)
library(tidyverse)
library(flextable)
library(stringr)
library(sf)
library(here)
library(tmap)
library(lfe)
library(modelsummary)
library(patchwork)
library(kableExtra)
```

@@

```{r run-analys, cache = TRUE, include = FALSE}
knitr::purl(
  here::here("GitControlled/Codes/1_analysis.Rmd"),
  here::here("GitControlled/Codes/1_analysis.R"),
)
source(here::here("GitControlled/Codes/1_analysis.R"))

knitr::purl(
  here::here("GitControlled/Codes/2_figures_tables.Rmd"),
  here::here("GitControlled/Codes/2_figures_tables.R"),
)
source(here::here("GitControlled/Codes/2_figures_tables.R"))

knitr::purl(
  here::here("GitControlled/Codes/3_appendix.Rmd"),
  here::here("GitControlled/Codes/3_appendix.R"),
)
source(here::here("GitControlled/Codes/3_appendix.R"))
```

@@

Individual county and year are indicated by $i$ and $t$, respectively. In the model, $y_{i,t}$ is crop yield for county $i$ in year$t$. $C_j$ is a dummy variable that takes 1 if observation $i$ is under category $C_j$, 0 otherwise. The observations are grouped into four categories: dryland production, observations with saturated thickness falling under the $[0-33.3\%)$j, $[33.3-66.7\%)$, and $[66.7\% - 100\%]$ quantile ranges of saturated thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. $\sigma_j$ represents the category specific intercept. $WB_{i,t}$ is water deficit defined as total reference evapotranspiration less precipitation between May through September. Category-specific impact of water deficit on yield is represented by $\sum_{k}^K \beta_k\phi_{j,k}(WB_{i,t})$, where $\phi_{j,k}$ is $k$th spline basis function and $\beta_k$ is its coefficient. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner. The model can be interpreted as hierarchical generalized additive model, where each of saturated thickness level category has its own yield response function to drought (cite). To control for time-invariant  heterogeneity across counties and yearly shocks, county fixed effects ($\alpha_i$) and year fixed effects ($\phi_t$) are included. Standard error estimation is clustered by county.

The impact of saturated thickness on the share of irrigated acres is estimated using a fractional logit model under the generalize additive model framework in which the impact of saturated thickness, average water deficit, and their interactions are estimated semi-parametrically, allowing for flexible representation of their impacts. The estimating equation is as follows:

@@

Figure \@ref(fig:geo-focus) shows the geographical coverage of the data for regression analysis for corn and soybean.

```{r, fig.dim = c(6, 5)}
#| label: geo-focus
#| fig-cap: "Geographic focus of the study for corn and soybean"

tm_map
```

First, county-level irrigated and dryland yield data for corn and soybean for 1985 through 2016 were obtained from the USDA NASS QuickStat. Both dryland and irrigated yield observations are used for counties that shares more than 75% of their area with the Ogallala aquifer. Only dryland yield observations are used for counties that do not satisfy the condition. We use dryland yield data outside of the Ogallala aquifer to supplement dryland yield data for more accurate estimation of the impact of water deficit for dryland production. Daily precipitation and reference ET were obtained from gridMET [@Abatzoglou2013]. Water deficit ($WB_{i,t}$) was then calculated by subtracting total precipitation from total reference evapotranspiration from May through September. Saturated thickness data were obtained from Dr. Erin Haacker [@haacker2016water]. Observations that have saturated thickness value lower than 12 meter were eliminated as saturated thickness below that is considered not suitable for irrigation (Tim citations here).

Figure \@ref(fig:deficit-hist) presents annual water deficit by year for the corn and soybean data. Water deficit is highly variable year to year. Year 2012 in particular stands out as an extremely dry year in the region. We use the year-to-year variations in water deficit to identify its impact on crop yield. Figure \@ref(fig:yield-hist) compares historical corn and soybean yields for dryland and irrigated production. The power of irrigation is clearly evident in this figure. First, irrigation provides protection against severe drought years like 2000, 2002, 2012. While the dryland yield dives down in those years, irrigated yields is almost unaffected by the severe droughts. First key objective of our study is to see if and how much the ability of irrigation is diminished when saturated thickness goes down. Figure \@ref(fig:sat-dist) presents the distribution of saturated thickness. Saturated thickness vary substantially, but most of the sites have saturated thickness of less than 50 meters. We exploit the observed variations in saturated thickness to identify its influence on the ability of irrigation to drought. Second, irrigated yield is always about 50% or higher than dryalnd yield irrespective of the degree of water deficit. We will investigate the negative impact of a decline in saturated thickness on the percentage of irrigated fields, thus reducing crop yields.

```{r }
#| fig-cap: "Distribution of water deficit for corn and soybean datasets"
#| label: deficit-hist
#| fig-width: 6
#| fig-height: 3.5

@@

```{r}
#| fig-cap: "Historical dryland and irrigated yields for corn and soybean"
#| label: yield-hist
#| fig-width: 6
#| fig-height: 3.5

@@

```{r}
#| fig-cap: "Distribution of saturated thickness"
#| label: sat-dist
#| fig-width: 6
#| fig-height: 3.5

@@

Regression results are presented graphically rather than in tables because individual coefficients estimates on spline basis functions cannot be readily interpreted and also because visualized results are much easier to interpret. Figure \@ref(fig:yield-response) shows yield response to water deficit for dryland production and groundwater-irrigated production at different levels of saturated thickness categories for corn (1st row) and soybean (2nd row).

```{r}
#| label: yield-response
#| fig-cap: "The impact of water deficit on dryland and irrigated yield by crop"
#| fig-width: 6
#| fig-height: 4

@@

In a fairly wet year without much water stress, dryland and irrigated production result in very similar crop yields for both soybean and corn yields. Irrigated yields are only slightly higher than dryland yields at a negative value of water deficit (precipitation is greater than reference evapotranspiration). However, dryland yield declines sharply and steadily as water deficit increases. On the contrary, irrigated corn yield actually increase as water deficit increases up to about water deficit of 400mm at all the saturated thickness levels for both corn and soybean. This is because irrigation can turn unfavorable high-heat conditions into favorable growing conditions up to a point (Tim, citations?). However, yield response to water deficit departs after deficit of about 400mm depending on the level of saturated thickness. While higher saturated thickness categories can withstand water deficit and crop yield stays about the same for higher water deficit values, the low saturated thickness category experience non-trivial declines in crop yields as water deficit increases.

@@

```{r ir-share-plot}
#| label: ir-share
#| fig-cap: "The impact of saturated thickness on the share of irriagate production by crop "
#| fig-width: 6
#| fig-height: 3.5

@@

As shown in \@ref(fig:ir-share), saturated thickness has significant impacts on the share of irrigated production for corn. As saturated thickness declines, the share of irrigated acres for corn decreases. As saturated thickness declines, more land will be under dryland production, which is more vulnerable to drought, damaging average corn yield. Unlike corn, a decline in saturated thickness does not have as large negative impact on the share of irrigated production for soybean as corn. This is likely because soybean is less susceptible to drought compared to corn.

@@

Taking into account both impacts of saturated thickness, figure \@ref(fig:tot-impact) shows how water deficit affects yield at different saturated thickness levels evaluated. For corn, in years with water deficit of 400mm or less saturated thickness does not have economically significant negative impacts because the difference in irrigated and dryland yields are small and also because irrigated yields are similar irrespective of the level of saturated thickness. The same can be said for soybean except that the soybean is more tolerant than corn and the saturated thickness starts to make differences at water deficit of 500mm of higher. In the year with water deficit of 800mm, average corn yield at saturated thickness of 186 meters is about 15% and 25% higher than at 75.7 meters and 29.6 meters, respectively. These findings illustrate that vulnerability to drought is heterogeneous among irrigated production depending on the level of saturated thickness. In summary, our studies confirms that a decline in saturated thickness makes yield production much more vulnerable to severe drought, and its negative impact is non-negligible.

It is also noteworthy that the impact of saturated thickness on crop yield in drought years is non-linear. Specifically, according to our estimates, the decline of saturated thickness from 186 meters to 75 meters would cause about one tonne/ha of damage in corn yield at water deficit of 1100mm. However, the decline of saturated thickness from 75 meters to 30 meters would cause about 1.5 tonne/ha of damage in corn yield. This means that the negative impact of a meter of decline in saturated thickness becomes larger as the aquifer depletion progresses. This has an important policy implication. Suppose we intend to limit groundwater pumping to maintain the stable level of groundwater stock by matching the natural replenishment rate for the purpose of sustainable groundwater-irrigated production. Then, we need to act early before saturated thickness becomes too low so that we do not need to product at a less productive steady state. The longer you wait, the harsher damage is going to be when you finally decide to place a policy that limits groundwater pumping.

```{r total-impact}
#| fig-cap: "The impact of water deficit on average crop yield by sturated thickness"
#| label: tot-impact
#| fig-width: 6
#| fig-height: 3.5
g_total
```

@@

\clearpage

# (APPENDIX) Appendix {-}

# Use SPEI in place of water deficit

Some studies use SPEI to represent drought severity. We also considered using SPEI, but the regressions with water deficit provides a better fit for both corn and soybean (see \@ref(tab:fit-reg)).

```{r}
#| label: fit-reg
#| tbl-cap: Comparison of R-squared of the regressions with water deficit and SPEI for corn and soybean

data.table(
  Crop = c("Corn", "Corn", "Soybean", "Soybean"),
  Variable = c("Water Deficit", "SPEI", "Water Deficit", "SPEI"),
  `R-squared` = c(r2_corn_wb, r2_corn_spei, r2_soy_wb, r2_soy_spei)
) %>%
  .[, `R-squared` := round(`R-squared`, digits = 3)] %>%
  dcast(Crop ~ Variable, value.var = "R-squared") %>%
  kbl(caption = "Comparison of R-squared of the regressions with water deficit and SPEI for corn and soybean", booktabs = T) %>%
  kable_styling(latex_options = c("hold_position"))
```

Qualitatively, the results with SPEI are similar to the regression results with water deficit as can be seen in \@ref(fig:spei).

```{r}
#| label: spei
#| fig-cap: The impact of SPEI on crop yield for corn and soybean

@@

\clearpage

>>>>

---
title: The Impacts of Aquifer Depletion on the Ability of Groundwater Irrigation to Buffer Against Undesirable Weather
authors:
- name: Taro Mieno
  address: Department of Agricultural Economics, University of Nebraska Lincoln
  email: tmieno2@unl.edu
  footnote: Something here
- name: Timothy Foster
  address: Department of WWW, University of Manchester
  email: author@xxx.edu
- name: Shunkei Kakimoto
  address: Department of Agricultural Economics, University of Nebraska Lincoln
  email: author@zzz.edu
- name: Nicholas Brozovic
  address: Department of Agricultural Economics, University of Nebraska Lincoln
  email: author@zzz.edu
abstract: |
  Abstract goes here. It should be a single paragraph of around 200 words.
acknowledgments: |
  We wish to thank...
bibliography: DRA.bib
csl: iopart.csl
format:
  pdf:
    cite-method: natbib
    keep-tex: true
---

@@

#--- packages ---#
library(data.table)
library(tidyverse)
library(flextable)
library(stringr)
library(sf)
library(here)
library(tmap)
library(lfe)
library(modelsummary)
library(patchwork)
```

@@

```{r run-analys}
#| include: false
#| cache: false
knitr::purl(
  here::here("Codes/1_analysis.Rmd"),
  here::here("Codes/1_analysis.R"),
)
source(here::here("Codes/1_analysis.R"))

knitr::purl(
  here::here("Codes/2_figures_tables.Rmd"),
  here::here("Codes/2_figures_tables.R"),
)
source(here::here("Codes/2_figures_tables.R"))

knitr::purl(
  here::here("Codes/3_appendix.Rmd"),
  here::here("Codes/3_appendix.R"),
)
source(here::here("Codes/3_appendix.R"))
```

@@

Individual county and year are indicated by $i$ and $t$, respectively. In the model, $y_{i,t}$ is crop yield for county $i$ in year$t$. $C_j$ is a dummy variable that takes 1 if observation $i$ is under catecotry $C_j$, 0 otherwise. The observations are grouped into four categories: dryland production, observations with saturated thickness falling under the $[0-33.3\%)$j, $[33.3-66.7\%)$, and $[66.7\% - 100\%]$ quantile ranges of saturated thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. $\sigma_j$ represents the category specific intercept. $WB_{i,t}$ is water deficit defined as total reference evapotranspitration less precipitation between May through September. Category-specific impact of water deficit on yield is represented by $\sum_{k}^K \beta_k\phi_{j,k}(WB_{i,t})$, where $\phi_{j,k}$ is $k$th spline basis function and $\beta_k$ is its coefficient. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner. The model can be interpreted as hierarchical generalized additive model, where each of saturated thickness level category has its own yield response function to drought (cite). To control for time-invariant  heterogeneity across counties and yearly shocks, county fixed effects ($\alpha_i$) and year fixed effects ($\phi_t$) are included. Standard error estimation is clustered by county.

The impact of saturated thickness on the share of irrigated acres is estiamted using a fractional logit model under the generalize additive model framwork in which the impact of saturated thickness, average water deficit, and their interactions are estimated semi-parametrically, allowing for flexible representation of their impacts. The estimating equation is as follows:

@@

@fig-geo-focus shows the geographical coverage of the data for regression analysis for corn and soybean.

```{r}
#| fig-cap: "Geographic focus of the study for corn and soybean"
#| label: fig-geo-focus
#| fig-width: 5
tm_map
```

First, county-level irrigated and dryland yield data for corn and soybean for 1985 through 2016 were obtained from the USDA NASS QuickStat. Both dryland and irrigated yield observations are used for counties that shares more than 75% of their area with the Ogallala aquifer. Only dryland yield observations are used for counties that do not satisfy the condition. We use dryland yield data outside of the Ogallala aquifer to supplement dryland yield data for more accurate estimation of the impact of water deficit for dryland production. Daily precipitation and reference ET were obtained from gridMET [@Abatzoglou2013]. Water deficit ($WB_{i,t}$) was then calculated by subtracting total precipitation from total reference evapotranspiration from May through September. Saturated thickness data were obtained from Dr. Erin Haacker [@haacker2016water]. Observations that have saturated thickness value lower than 12 meter were eliminated as saturated thicknees below that is considered not suitable for irrigation (Tim citations here).

@fig-deficit-hist presents annual water deficit by year for the corn and soybean data. Water deficit is highly variable year to year. Year 2012 in particular stands out as an extremely dry year in the region. We use the year-to-year variations in water deficit to idenfity its impact on crop yield. @fig-yield-hist compares historical corn and soybean yields for dryland and irrigated production. The power of irrigation is clearly evident in this figure. First, irrigation provides protection against severe drought years like 2000, 2002, 2012. While the dryland yield dives down in those years, irrigated yields is almost unaffected by the severe droughts. First key objective of our study is to see if and how much the ability of irrigation is dminisihed when saturated thickness goes down. @fig-sat-dist presents the distribution of saturated thickness. Saturated thickness vary substantially, but most of the sites have saturated thickness of less than 50 meteres. We exploit the observed variations in saturated thickness to identify its influence on the ability of irrigation to drought. Second, irrigated yield is always about 50% or higher than dryalnd yield irrespective of the degree of water deficit. We will investigate the negative impact of a decline in saturated thickness on the percentage of irrigated fields, thus reducing crop yields.

```{r}
#| fig-cap: "Distribution of water deficit for corn and soybean datasets"
#| label: fig-deficit-hist
#| fig-width: 5

@@

```{r}
#| fig-cap: "Historical dryland and irrigated yields for corn and soybean"
#| label: fig-yield-hist
#| fig-width: 5

@@

```{r}
#| fig-cap: "Distribution of saturated thickness"
#| label: fig-sat-dist
#| fig-width: 5

@@

Regression results are presented graphically rather than in tables because individual coefficients estimates on spline basis functions cannot be readily interpreted and also because visualized results are much easier to interpret. Figure @fig-yield shows yield response to water deficit for dryland production and groundwater-irrigated production at different levels of saturated thickness categories for corn (1st row) and soybean (2nd row).

```{r}
#| label: fig-yield
#| fig-cap: "The impact of water deficit on dryland and irrigated yield by crop"
#| fig-width: 5

@@

In a fairly wet year without much water stress, dryland and irrigated production result in very similar crop yields for both soybean and corn yields. Irrigated yields are only slighly higher than dryland yields at a negative value of water deficit (precipitation is greater than reference evapotraspiration). However, dryland yield declines sharply and steadily as water deficit increases. On the contrary, irrigated corn yield actually increase as water deficit increases up to about water deficit of 400mm at all the saturated thickness levels for both corn and soybean. This is because irrigation can turn unfavorable high-heat conditions into favorable growing conditions up to a point (Tim, citations?). However, yield response to water deficit departs after deficit of about 400mm depending on the level of saturated thickness. While higher saturated thickness categories can withstand water deficit and crop yield stays about the same for higer watger deficit values, the low saturated thickness category experience non-trivial declines in crop yields as water deficit increases.

@@

```{r ir-share-plot}
#| label: fig-ir-share
#| fig-cap: "The impact of saturated thickness on the share of irriagate production by crop "

@@

As shown in @fig-ir-share, saturated thickness has significant impacts on the share of irrigated production for corn. As saturated thickness declines, the share of irrigated acres for corn decreases. As saturated thickness declines, more land will be under dryland production, which is more vulnerable to drought, damaging average corn yield. Unlike corn, a decline in saturated thickness does not have as large negative impact on the share of irrigated production for soybean as corn. This is likely because soybean is less susceptible to drought compared to corn.

@@

Taking into account both impacts of saturated thickness, figure @fig-tot-impact shows how water deficit affects yield at different saturated thickness levels evaluated. For corn, in years with water deficit of 400mm or less saturated thickness does not have economically significant negative impacts because the difference in irrigated and dryland yields are small and also because irrigated yields are similar irrespective of the level of saturated thickness. The same can be said for soybean except that the soybean is more tolerant than corn and the saturated thickness starts to make differences at water deficit of 500mm of higher. In the year with water deficit of 800mm, average corn yield at saturated thickness of 186 meters is about 15% and 25% higher than at 75.7 meters and 29.6 meters, respectively. These findings illustrate that vulnerability to drought is heterogeneous among irrigated production depending on the level of saturated thickness. In summary, our studies confirms that a decline in saturated thickness makes yield production much more vulnerable to severe drought, and its negative impact is non-negligible.

It is also noteworthy that the impact of saturated thickness on crop yield in drought years is non-linear. Specifically, according to our estimates, the decline of saturated thickness from 186 meters to 75 meters would cause about one tonne/ha of damage in corn yield at water deficit of 1100mm. However, the decline of saturated thickness from 75 meters to 30 meters would cause about 1.5 tonne/ha of damage in corn yield. This means that the negative impact of a meter of decline in saturated thickness becomes larger as the aquifer depletion progresses. This has an important policy implication. Suppose we intend to limit groundwater pumping to maintain the stable level of groundwater stock by matching the natural replenishment rate for the purpose of sustainable groundwawter-irrigated production. Then, we need to act early before saturated thickness becomes too low so that we do not need to product at a less productive steady state. The longer you wait, the harsher damage is going to be when you finally decide to place a policy that limits groundwater pumping.

```{r total-impact}
#| fig-cap: "The impact of water deficit on average crop yield by sturated thickness"
#| label: fig-tot-impact
g_total
```

@@

\newpage

# Appendix {.appendix}

## Use SPEI in place of water deficit

Some studies use SPEI to represent drought severity. We also considered using SPEI, but the regressions with water deficit provides a better fit for both corn and soybean (see @tbl-fit).

```{r}
#| label: tbl-fit
#| tbl-cap: Comparison of R-squared of the regressions with water deficit and SPEI for corn and soybean

data.table(
  Crop = c("Corn", "Corn", "Soybean", "Soybean"),
  Variable = c("Water Deficit", "SPEI", "Water Deficit", "SPEI"),
  `R-squared` = c(r2_corn_wb, r2_corn_spei, r2_corn_wb, r2_soy_spei)
) %>%
modelsummary::datasummary(`R-squared` ~ Crop * Variable * (Mean), data = .)
```

Qualitatively, the results with SPEI are similar to the regression reulsts with water deficit as can be seen in @fig-spei.

```{r}
#| label: fig-spei
#| fig-cap: The impact of SPEI on crop yield for corn and soybean

@@

<!-- ## Continuous saturated thickness -->

\newpage

### References

::: {#refs}
:::
